---
title: "Using fledge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This guide will demonstrate how to use {fledge}, using a mock R package as an example.
We are going to call our package "{SuperFrob}".
We will develop it from scratch and also maintain a changelog as the development progresses.
Finally, we will demonstrate how this changelog can eventually be converted to release notes when the package is submitted to CRAN.

This guide shows mostly R code, and uses Bash shell code for Git commands.

```{r setup, include = FALSE}
if (identical(Sys.getenv("IN_PKGDOWN"), "true")) {
  options(crayon.enabled = TRUE)
}
knitr::opts_chunk$set(
  collapse = TRUE
)
Sys.setenv(GIT_AUTHOR_DATE = "2020-12-12 12:47:37Z")
Sys.setenv(GIT_AUTHOR_NAME = "Kirill Müller")
Sys.setenv(GIT_AUTHOR_EMAIL = "krlmlr+r@mailbox.org")
Sys.setenv(GIT_COMMITTER_DATE = "2020-12-12 12:47:37Z")
Sys.setenv(GIT_COMMITTER_NAME = "Kirill Müller")
Sys.setenv(GIT_COMMITTER_EMAIL = "krlmlr+r@mailbox.org")
```

## Set up the development environment

Before we start development for {SuperFrob}, we set up the basic development environment required for any typical R package.

### Create a package

We will start by creating a new package.
For this demo, the package is created in a temporary directory.
A real project will live somewhere in your home directory.

```{r tempdir}
tempdir <- file.path(tempdir(), "fledge")
dir.create(tempdir)
```

The `usethis::create_package()` function sets up a package project ready for development.
The output shows the details of the package created.

```{r create-package}
pkg <- usethis::create_package(file.path(tempdir, "SuperFrob"))
```

In an interactive RStudio session, a new window opens.
Users of other environments would change the working directory manually.
For this demo, we manually set the active project.

```{r set-focus, include = FALSE}
wd <- getwd()

knitr::knit_hooks$set(
  pkg = function(before, options, envir) {
    if (before) {
      wd <<- setwd(pkg)
    } else {
      setwd(wd)
    }

    invisible()
  }
)

knitr::opts_chunk$set(pkg = TRUE)
```

```{r pkg-location}
usethis::proj_set()
```

The infrastructure files and directories that comprise a minimal R package are created:

```{r dir-tree}
fs::dir_tree()
```


### Create and configure a Git repository

Next we set up this package for development and create a Git repository for the package.
We use `git init` to get a predictable branch name.

```{bash git, error = (.Platform$OS.type == "windows")}
git init --initial-branch=main
```

```{r git-config, include = FALSE}
# ...but only in interactive mode.
git2r::config(user.name = "Test", user.email = "test@user.org")
```

In interactive mode, the `usethis::use_git()` function creates an initial commit, and the repository is in a clean state.
In the demo this needs to be carried out manually.

```{bash git-initial-commit, error = (.Platform$OS.type == "windows")}
git add .
git commit -m "Initial commit."
```

```{bash git-show-init, error = (.Platform$OS.type == "windows")}
git log --numstat
git status
```

For working in branches, it is recommended to turn off fast-forward merging:

```{bash git-no-ff, error = (.Platform$OS.type == "windows")}
git config merge.ff false
```

An alternative is to use squash commits.

### Create initial NEWS.md file

An initial NEWS file can be created with `usethis::use_news_md()`.

```{r init-news}
usethis::use_news_md()
```

Let's take a look at the contents:

```{r init-news-review}
news <- readLines(usethis::proj_path("NEWS.md"))
cat(news, sep = "\n")
```

This file needs to be tracked by Git:

```{bash init-news-commit, error = (.Platform$OS.type == "windows")}
git add .
git status
git commit -m "Initial NEWS.md ."
```

## The development phase

### Create an R file

Now we start coding in the functionality for the package.
We start by creating the new R file called `super.R` and adding frobnication code.

```{r use-super}
usethis::use_r("super")
writeLines("# frobnicate", "R/super.R")
```

We commit this file to Git with a relevant message.
Note the use of the bullet (-) before the commit message.
This indicates that the message should be included in `NEWS.md` when it is updated.

```{bash commit-super, error = (.Platform$OS.type == "windows")}
git add .
git commit -m "- Add support for frobmication."
```

### Create a test

The code in `super.R` warrants a test:

```{r use-super-test}
usethis::use_test("super")
cat(readLines("tests/testthat/test-super.R"), sep = "\n")
```

In a real project we would substitute the testing code from the template by real tests.
In this demo we commit straight away, again with a bulleted message.

```{bash commit-super-test, error = (.Platform$OS.type == "windows")}
git add .
git status
git commit -m "- Add tests for frobnication."
```

### Update NEWS.md

We use `fledge::bump_version()` to assign a new dev version number to the package and also update `NEWS.md`.

```{r bump}
fledge::bump_version()
```

### Review NEWS.md

Let us see what `NEWS.md` looks like after that bump.

```{r news-review}
news <- readLines("NEWS.md")
cat(news, sep = "\n")
```

While reviewing we notice that there was a typo in one of the comments.
Let's fix the typo:

```{r news-tweak}
news <- gsub("frobmication", "frobnication", news)
cat(news, sep = "\n")
writeLines(news, "NEWS.md")
```

This does not affect the original commit message, only `NEWS.md`.

### Finalize version

After tweaking `NEWS.md`, it is important to use `fledge::finalize_version()` and not to commit manually.
This ensures that the tag is set to the correct version in spite of the `NEWS.md` update.
It should be called when `NEWS.md` is manually updated.

```{r news-finalize}
fledge::finalize_version()
```

When done, we push to the *remote* (GitHub or another Git service) with tags using `fledge::finalize_version(push = TRUE)`, as advised.
In this demo, no remote repository is set up, and the push process is not shown.

### Change code and commit

{SuperFrob} with frobnication is tested, now we want to enhance with super-frob.
This requires changes to the code, and perhaps a new test.
We create a branch and switch to this branch to implement this.

```{bash super-frob-branch, error = (.Platform$OS.type == "windows")}
git checkout -b f-super-frob
```

On the branch, separate commits are used for the tests and the implementation.
These commit messages do not need to be formatted specially, because {fledge} will ignore them anyway.

```{r super-frob}
usethis::use_test("frob")
```


```{bash super-frob-commit, error = (.Platform$OS.type == "windows")}
git add .
git status
git commit -m "Add super-frob tests."
```


```{r super-frob-2}
usethis::use_r("frob")
writeLines("# super-frob", "R/frob.R")
```


```{bash super-frob-2-commit, error = (.Platform$OS.type == "windows")}
git add .
git status
git commit -m "Add super-frob implementation."
```

This branch can be pushed to the remote as usual.
Only when merging we specify the message to be included in the changelog, again with a bullet.[^merge-ff]

[^merge-ff]: Note that we really need a merge commit here; the default is to fast-forward which doesn't give us the opportunity to insert the message intended for the changelog.
Earlier, we set the `merge.ff` config option to `"false"` to achieve this.

```{bash super-frob-merge, error = (.Platform$OS.type == "windows")}
git checkout main
git merge f-super-frob --no-commit
git status
git commit -m "- Super-frobnication enabled."
```

The same strategy can be used when merging a pull/merge/... request on GitHub, GitLab, ...: use bullet points in the merge commit message to indicate the items to include in `NEWS.md`.

Now that we have added super-frobnication support to our package, it is time to bump the version again.

```{r bump-two}
fledge::bump_version()
news <- readLines("NEWS.md")
cat(news, sep = "\n")
```

## Prepare for release

After multiple cycles of development, review and testing, we decide that our package is ready for release to CRAN.
This is where {fledge} simplifies the release process by making use of the all the commit messages that we provided earlier.

### Bump version for release

We wish to release this package as a patch and so we use `fledge::bump_version()` with the "patch" argument.

```{r bump-patch}
fledge::bump_version("patch")
```

This updates the version of our package to `r desc::desc_get_version()`.

### Generate release notes

We review the `NEWS.md` that were generated by {fledge}:

```{r patch-news-review}
news <- readLines("NEWS.md")
cat(news, sep = "\n")
```

Some of the intermediate commit messages are not relevant in the release notes for this release.
We need to edit `NEWS.md` to convert the changelog to meaningful release notes.

```{r patch-news-edit}
length(news) <- 5
news[3:5] <- c(
  "Initial CRAN release.",
  "",
  "Basic functionality: Super-frobnication."
)
cat(news, sep = "\n")
writeLines(news, "NEWS.md")
```

Unlike with development versions, we commit the changes to `NEWS.md` manually:

```{bash patch-news-write, error = (.Platform$OS.type == "windows")}
git add .
git status
git commit -m "Update NEWS."
```

The package is now ready to be released to CRAN.
I prefer `devtools::use_release_issue()` to create a checklist of things to do before release, and `devtools::submit_cran()` to submit.
The `devtools::release()` function is a more verbose alternative.

## After release

Some time passed and our {SuperFrob} package was accepted on CRAN.
At this stage, {fledge} can help to tag the released version and create a new version for development.

### Tag version

It is now the time to tag the released version using the `fledge::tag_version()` function.

```{r tag}
fledge::tag_version()
```

It is advised to push to remote, with `git push --tags` from the command line, or your favorite Git client.

### Create GitHub release

If your package is hosted on GitHub, `usethis::use_github_release()` creates a draft GitHub release from the contents already in `NEWS.md`.
You need to submit the draft release from the GitHub release page.

### Restart development

We will now make the package ready for future development.
The `fledge::bump_version()` takes care of it.

```{r bump-dev}
fledge::bump_version()
news <- readLines("NEWS.md")
cat(news, sep = "\n")
```

Push to remote.
