---
title: "Using Fledge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This guide will demonstrate how to use `fledge` functionality with a mock R package.
We are going to call our package "`SuperFrob`".
We will develop it from scratch and also maintain a change-log as the development progresses.
Finally, we will demonstrate how this change log can eventually be converted to release notes when the package is submitted to CRAN.

```{r setup, include = FALSE}
options(crayon.enabled = TRUE)
fansi::set_knit_hooks(knitr::knit_hooks, "output")
fansi::set_knit_hooks(knitr::knit_hooks, "message")
```

## Set up the development environment

Before we start development for "`SuperFrob`", we set up the basic development environment required for any typical R package.

### Create a Package

We will start by creating a new package.
For this demo, the package is created in a temporary directory.
A real project will live somewhere in your home directory.

```{r tempdir}
tempdir <- tempfile("fledge")
dir.create(tempdir)
```

The `usethis::create_package()` function sets up a package project ready for development.
The output shows the details of the package created.

```{r create-package}
pkg <- usethis::create_package(file.path(tempdir, "SuperFrob"))
```

In an interactive RStudio session, a new window opens.
Users of other environments would change the working directory manually.
For this demo, we manually set the active project.

```{r set-focus, include = FALSE}
knitr::knit_hooks$set(
  pkg = function(before, options, envir) {
    if (before) {
      setwd(pkg)
    }
  }
)
```

```{r pkg-location, pkg = TRUE}
usethis::proj_set(pkg)
```

The infrastructure files and directories that comprise a minimal R package are created:

```{r dir-tree}
fs::dir_tree()
```


### Create and configure a Git repository

Next we set up this package for development and create a Git repository for the package.

```{r git}
usethis::use_git()
```

The `usethis::use_git()` function creates an initial commit, the repository is in a clean state.

```{r git-config, include = FALSE, eval = !interactive()}
# ...but only in interactive mode.
git2r::config(user.name = "Test", user.email = "test@user.org")
git2r::add(path = ".")
git2r::commit(message = "Initial commit.")
```

```{r git-show-init}
git2r::commits()
git2r::status()
```

### Create initial NEWS.md file

An initial NEWS file can be created with `usethis::use_news_md()`.

```{r init-news}
usethis::use_news_md()
```

Let's take a look at the contents:

```{r init-news-review}
news <- readLines(usethis::proj_path("NEWS.md"))
cat(news, sep = "\n")
```

This file needs to be tracked by Git:

```{r init-news-commit}
git2r::add(path = ".")
git2r::status()
git2r::commit(message = "Initial NEWS.md .")
```

## The development phase

### Create an R file

Now we start coding in the functionality for the package.
We start by creating the new R file called `super.R` and adding frobnication code.

```{r use-super}
usethis::use_r("super")
writeLines("# frobnicate", "R/super.R")
```

We commit this file to Git with a relevant message.
Note the use of the bullet (-) before the commit message.
This indicates that the message should be included in `NEWS.md` when it is updated.

```{r commit-super}
git2r::add(path = ".")
git2r::commit(message = "- Add support for frobmication.")
```

### Create a test

The code in `super.R` warrants a test:

```{r use-super-test}
usethis::use_test("super")
cat("tests/testthat/test-super.R", sep = "\n")
```

In a real project we would substitute the testing code from the template by real tests.
In this demo we commit straight away, again with a bulleted message.

```{r commit-super-test}
git2r::add(path = ".")
git2r::status()
git2r::commit(message = "- Add tests for frobnication.")
```

### Update NEWS.md

We use `fledge::bump_version()` to assign a new dev version number to the package and also update `NEWS.md`.

```{r bump}
fledge::bump_version()
```

### Review NEWS.md

Let us see what `NEWS.md` looks like after that bump.

```{r news-review}
news <- readLines("NEWS.md")
cat(news, sep = "\n")
```

While reviewing we notice that there was a typo in one of the comments.
Let's fix the typo:

```{r news-tweak}
news <- gsub("frobmication", "frobnication", news)
cat(news, sep = "\n")
writeLines(news, "NEWS.md")
```

This does not affect the original commit message, only `NEWS.md`.

### Finalize Version

After tweaking `NEWS.md`, it is important to use `fledge::finalize_version()` and not to commit manually.
This ensures that the tag is set to the correct version in spite of the `NEWS.md` update.
It should be called every time `NEWS.md` is manually updated.

```{r news-finalize}
fledge::finalize_version()
```

When done, we push to the *remote* (GitHub or another Git service) with tags, using `git push --tags` from the command line or your favorite Git client.
In this demo, no remote repository is set up, and the push process is not shown.

### Change code and commit

SuperFrob with frobnication is tested, now we want to enhance with super-frob.
This requires changes to the code, and perhaps a new test.
We create a branch and switch to this branch to implement this.

```{r super-frob-branch}
git2r::branch_create(git2r::last_commit(), name = "f-super-frob")
git2r::checkout(branch = "f-super-frob")
```

On the branch, separate commits are used for the tests and the implementation.
These commit messages do not need to be formatted specially, because {fledge} will ignore them anyway.

```{r super-frob}
usethis::use_test("frob")
git2r::add(path = ".")
git2r::status()
git2r::commit(message = "Add super-frob tests.")

usethis::use_r("frob")
writeLines("# super-frob", "R/frob.R")
git2r::add(path = ".")
git2r::status()
git2r::commit(message = "Add super-frob implementation.")
```

This branch can be pushed to the remote as usual.
Only when merging we specify the message to be included in the changelog, again with a bullet.

```{r super-frob-merge}
git2r::checkout(branch = "master")
git2r::merge(".", "f-super-frob", commit_on_success = FALSE)
git2r::status()
git2r::commit(message = "- Super-frobnication enabled.")
```

Now that we have added super-frobnication support to our package, it is time to bump the version again.

```{r bump-two}
fledge::bump_version()
news <- readLines("NEWS.md")
cat(news, sep = "\n")
```

## Prepare for release

After multiple cycles of development, review and testing, we decide that our package is ready for release to CRAN.
This is where `fledge` simplifies the release process by making use of the all the commit messages that we provided earlier.

### Fledge for release 

We wish to release this package as a patch and so we use `fledge::bump_version()` with the "patch" argument.

```{r bump-patch}
fledge::bump_version("patch")
```

This updates the version of our package to `r desc::desc_get_version()`.

### Generate release notes

We review the `NEWS.md` that were generated by `Fledge`:

```{r patch-news-review}
news <- readLines("NEWS.md")
cat(news, sep = "\n")
```

Some of the intermediate commit messages are not relevant in the release notes for this release.
We need to edit `NEWS.md` to convert the changelog to meaningful release notes.

```{r patch-news-edit}
length(news) <- 5
news[3:5] <- c(
  "Initial CRAN release.",
  "",
  "Basic functionality: Super-frobnication."
)
cat(news, sep = "\n")
writeLines(news, "NEWS.md")
```

Unlike with development versions, we commit the changes to `NEWS.md` manually:

```{r patch-news-write}
git2r::add(path = ".")
git2r::status()
git2r::commit(message = "Update NEWS.")
```

The package is now ready to be released to CRAN.
I prefer `devtools::use_release_issue()` to create a checklist of things to do before release, and `devtools::submit_cran()` to submit.
The `devtools::release()` function is a more verbose alternative.

## After release

Some time passed and our SuperFrob package was accepted on CRAN.
At this stage, `Fledge` can help to tag the released version and create a new version for development.

### Tag version

It is now the time to tag the released version using the `fledge::tag_version()` function.

```{r tag}
fledge::tag_version()
```

It is advised to push to remote, with `git push --tags` from the command line, or your favorite Git client.

### Create GitHub release

If your package is hosted on GitHub, `usethis::use_github_release()` creates a draft GitHub release from the contents already in `NEWS.md`.
You need to submit the draft release from the GitHub release page.

### Restart development

We will now make the package ready for future development.
The `fledge::bump_version()` takes care of it.

```{r bump-dev}
fledge::bump_version()
news <- readLines("NEWS.md")
cat(news, sep = "\n")
```

Push to remote.
